---
title: "ResidentialCommercialDev"
format: html
execute: 
  freeze: auto
---

Menace 1: Développement résidentiel et commercial

None of the BD layers correspond to Menace 1. I think that pulling buildings from OSM maps is a good strategy for this one.

```{r OSM buildings}
#install.packages("osmextract")
library(osmextract)
library(sf)
options(timeout=9999)
qc_buildings <- oe_get(
  place = "Quebec",
  provider = "geofabrik",
  layer = "multipolygons",
  query = "SELECT * FROM multipolygons WHERE building IS NOT NULL"
)

print(paste("Total buildings in OSM in Quebec:", length(st_geometry(qc_buildings)[[1]]) ))

## Projection for perserving area
qc_buildings_proj <- st_transform(qc_buildings, 32198)

## Area per building
areas <- st_area(qc_buildings_proj)

## Calculate total area (in km2)

total_area <- sum(areas, na.rm = TRUE) |> units::set_units(km2)
total_area # km2
```

```{r}

## OSM regions

qc_admin <- oe_get(
  place = "Quebec",
  provider = "geofabrik",
  layer = "multipolygons",
  query = "SELECT * FROM multipolygons WHERE boundary IS NOT NULL AND admin_level IS 6"
)

MRC <- qc_admin |>
  filter(!is.na(name)) |>
  st_make_valid()


qc_buildings <- st_transform(qc_buildings_proj, st_crs(regions))
centroids <- st_centroid(qc_buildings)

buildings_with_region <- st_join(centroids, MRC)


regions = st_read('RegionAdminQC/SHP/regio_l.shp')
qc_buildings_proj_region <- st_transform(qc_buildings_proj, st_crs(regions))

plot(regions)

library(dplyr)
sf::sf_use_s2(FALSE)
buildings_by_region <- st_join(qc_buildings_proj_region, regions)


```

```{r CanadaBuildingDatabase}
## Get building polygons
qc_1 = st_read("~/Maps/ODB_v3_QC_1/ODB_v3_QC_1.gpkg")
qc_2 = st_read("~/Maps/ODB_v3_QC_2/ODB_v3_QC_2.gpkg")

qc_centroid_1 = st_centroid(qc_1)
qc_centroid_2 = st_centroid(qc_2)

qc_centroid = rbind(qc_centroid_1, qc_centroid_2)
nrow(qc_centroid)

st_crs(qc_centroid)
coords <- st_coordinates(qc_centroid)

## Get Quebec's outer and internal boundaries

qc_admin <- oe_get(
  place = "Quebec",
  provider = "geofabrik",
  layer = "multipolygons",
  query = "SELECT * FROM multipolygons WHERE boundary IS NOT NULL AND admin_level IS 5"
)
qc_admin

qc_border <- st_transform(qc_admin, st_crs(qc_centroid))
qc_border_vect <- terra::vect(qc_border)

st_crs(qc_border)

## Convert to sp objects

library(sp)
pts <- SpatialPoints(coords, proj4string = CRS(st_crs(qc_centroid)$wkt))

library(spatstat.geom)
library(spatstat.explore)

# 
library(terra)

# convert points to SpatVector
pts <- vect(qc_centroid)

# raster template (adjust resolution!)
r <- rast(ext(pts), resolution = 5000)  # 5 km grid

# rasterize counts
r_count <- rasterize(pts, r, fun = "count", background = 0)

# smooth (Gaussian kernel)
kde <- focal(r_count, w = focalMat(r_count, 5000, type = "Gauss"))


# convert to spatstat
win <- as.owin(st_union(qc_border))

rx = callr::r_bg(function(win, coords) {
  spatstat.geom::ppp(
    x = coords[,1],
    y = coords[,2],
    window = win
  )
}, args = list(win, coords)
)
rx$is_alive()
pp <- rx$get_result()

kde <- density(pp, sigma = 500)

# library(spatialEco)
# kde = sf.kde(qc_centroid, bw=1000, res=10000)

## Add quebec boundary layer 


# terra::plot(st_geometry(qc_centroid_2), col = "red", pch = 16)
plot(kde)
terra::plot(qc_border_vect, add = TRUE, col = NA, border = "black", lwd = 1)


```

According to [2016 census data](https://www12.statcan.gc.ca/census-recensement/2016/as-sa/98-200-x/2016005/98-200-x2016005-fra.cfm), there are over 3.5 million private dwellings in Quebec, 45.4% of which are single-family homes.

In 2025, the governament of Canada released the [Open Database of Buildings](https://www150.statcan.gc.ca/n1/pub/34-26-0001/342600012018001-eng.htm) which contains 14.4 million records of buildings from across Canada, of which over 3.6 million are in Quebec. In contrast, open street maps, only has 1.6 million building polygons in Quebec.

Another way to measure this would be residential and commercial zoings. Whearas buildings would be an underestimation of the residential and commercial threat, zoning would be an overestimation. Unfortunately, there is not yet a dataset that includes zoning from all of the municipalities in Quebec (though one does exist for the CCM de Montreal, and several others are publically avaliable).

```{r}
#| fig.caption: Zoning of the city of Quebec. 

zonage = st_read("~/Maps/vdq-zonagemunicipalzones.geojson")
plot(zonage)
```
